---
title: "PSET_1"
author: "Suparna Kumar"
format: pdf
editor: visual
---

## How to apply the quadratic equation

This lesson demonstrates how to apply the quadratic equation with an example. We will also graphically demonstrate the solution.

### Applying quadratic equation

The roots of a quadratic equation of the form $f(x) = ax^2+bx+d=0$ are:

$$
x = \frac{-b \pm \sqrt{b^2-4ad}}{2a}
$$ In this example, we will only report the real roots.

```{r}
# Define the variables
a <- 1
b <- 3
d <- 2

# Calculating the roots
a <- as.complex(a)
roots <- c( ((-b + sqrt(b^2 - 4*a*d)) / (2*a)),
              ((-b - sqrt(b^2 - 4*a*d)) / (2*a)) )
if(all(Im(roots) == 0)){
  roots <- Re(roots)
  print(roots)
}
```

### Graphical solution of quadratic equation

The x-intercepts of the graph give the roots of the quadratic equation if the roots are real. Here, they are marked in red.

```{r}
#| echo: false
# Domain over which to plot f(x)
x <- seq(-5, 5,len = 100)

# Define f(x)
f <- a*x^2 + b*x + d

# Plot f(x)
plot(x, f, type = "l")
abline(h = 0) # x-axis
abline(v = 0) # y-axis
abline(v = roots, col = "red", lwd = 2) # roots
```
